---
title: "Customize your Risk Maps"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mappestRisk)
```

## Customize your risk map

The `map_risk()` function will usually serve to explore risk projections. However, the raw configuration of the plots in `terra::plot()` does not offer many possibilities to customize visualization for professional purposes. Since, the output of `map_risk()` that come across with the standard plot is a `SpatRaster`, this raster can be then exported using `terra::writeRaster` in *\*.tiff* format or it can be visualized using R packages for raster visualization such as `tidyterra` [@tidyterra] –under `ggplot2` [@ggplot2] logic– or `rasterVis` [@rasterVis], among others.

Here we provide an example of customization with `tidyterra` for the latest risk raster obtained in the [Generating Risk Maps](articles/generate-risk-maps.html) article.

```{r, message = FALSE, warning=FALSE, echo=FALSE, include=FALSE}

# STEP 1: Fit TPCs
data("aphid")
fitted_tpcs_aphid <- fit_devmodels(temp = aphid$temperature,
                                   dev_rate = aphid$rate_value,
                                   model_name = "all")
## optional: show predictions

# plot_devmodels(temp = aphid$temperature,
#                dev_rate = aphid$rate_value,
#                fitted_parameters = fitted_tpcs_aphid,
#                species = "Brachycaudus schwartzi",
#                life_stage = "nymphs")

# STEP 2: Obtain predictions and bootstrapped TPCs

preds_boots_aphid <- predict_curves(temp = aphid$temperature,          
                                    dev_rate = aphid$rate_value,
                                    fitted_parameters = fitted_tpcs_aphid,
                                    model_name_2boot = c("briere2", "lactin2"),
                                    propagate_uncertainty = TRUE,
                                    n_boots_samples = 100)
# optional: show simulated curves:
# plot_uncertainties(bootstrap_uncertainties_tpcs = preds_boots_aphid,
#                    temp = aphid$temperature,
#                    dev_rate = aphid$rate_value,
#                    species = "Brachycaudus schwartzi",
#                    life_stage = "Nymphs")

# STEP 3: Calculate thermal suitability boundaries

boundaries_aphid <- therm_suit_bounds(preds_tbl = preds_boots_aphid,       
                                      model_name = "lactin2",        
                                      suitability_threshold = 80) 
print(boundaries_aphid)

andalucia_sf <- mapSpain::esp_get_ccaa(ccaa = "Andalucía")

risk_rast_andalucia <- map_risk(t_vals = boundaries_aphid, 
                                path = tempdir(), # directory to download data 
                                region = andalucia_sf,    
                                mask = TRUE,
                                plot = TRUE,
                                interactive = FALSE,
                                verbose = TRUE)
```

. First, let's select the *risk* layer (if we don't specifically select it, the latest –sd– is used by default[^1]).

[^1]: Please, note that if the aim is to generate a custom map for risk uncertainty –i.e., the *uncertainty sd layer*, it must be selected and, accordingly, one may need to manually adjust the breaks of the color palettes and its related code.

```{r}
library(ggplot2)
risk_layer_andalucia <- risk_rast_andalucia |> 
  tidyterra::select("mean") # or alternatively, risk_layer_andalucia <- risk_rast_andalucia["mean"]
```

First, we will obtain an `sf` for the world map using `rnaturalearth` [@rnaturalearthpackage] and EPSG: 3035 as CRS to place underneath the risk map:

```{r}
worldmap_sf <- rnaturalearth::ne_countries(scale = 10,
                                           type = "countries",
                                           returnclass = "sf") |>
  sf::st_transform(25830) # <- from default EPSG: 4326 to target EPSG:3035
```

Then, we will crop it using the boundaries of our risk raster without `NAs`. However, we must project the raster to the target CRS (in this case: EPSG:3035)

```{r}
noNA_risk_rast <- risk_layer_andalucia |> 
  tidyterra::drop_na()
risk_rast_andalucia_epsg25830 <- terra::project(x = noNA_risk_rast,
                                               "EPSG:25830")
bbox_rast <- sf::st_bbox(risk_rast_andalucia_epsg25830)
```

Then, we will use a custom palette based on *magma* palette from `viridis` package with a custom color for `0`'s:

```{r}
my_risk_palette <- c("#AFC8C8", viridis::magma(12, direction = -1))
                      
```

For now, the map with these custom options and `tidyterra::geom_spatraster()` function will look as follows:

```{r}
max_risk_andalucia <- terra::minmax(risk_rast_andalucia_epsg25830)[2] # for manual scale fill visualization 

aphid_risk_map <- ggplot2::ggplot() +
  ggplot2::geom_sf(data = worldmap_sf, # <- countries background
                   fill = "gray85") +
  ggplot2::lims(x = c(bbox_rast[1], bbox_rast[3]),
                y = c(bbox_rast[2], bbox_rast[4])) +
  tidyterra::geom_spatraster(data = risk_rast_andalucia_epsg25830, 
                             maxcell = Inf) +
  ggplot2::geom_sf(data = worldmap_sf,
                   fill = NA,
                   color = "lightblue4") +
  ggplot2::scale_fill_gradientn(colours = my_risk_palette[1:(max_risk_andalucia + 1)], # to preserve the scales
                                na.value = "transparent",
                                breaks = seq(0, 12, by = 1))
print(aphid_risk_map)
```

We can improve the legend visualization using `ggplot2::guides()` is not correct, but we may correct it:

```{r}
aphid_risk_map <- aphid_risk_map +
  guides(fill = guide_legend(ncol = 13, #from 0 to 12 possible values
                             nrow = 1, 
                             byrow = TRUE)) +
  theme(legend.position = "bottom")
print(aphid_risk_map)
```

We can further improve visualization by changing the theme, adding titles and avoid borders overlapping with cells using bathymetry and adjacent region polygons:

```{r}
bath_spain <- mapSpain::esp_get_hypsobath() |> 
  sf::st_transform(25830) |> 
  dplyr::filter(val_sup <= 0)

andalucia_sf <- mapSpain::esp_get_ccaa("Andalucia") |> 
  sf::st_transform(25830)

#and borders:
clm_sf <- mapSpain::esp_get_ccaa("Castilla-La Mancha") |> 
  sf::st_transform(25830)

extremadura_sf <- mapSpain::esp_get_ccaa("Extremadura") |> 
  sf::st_transform(25830)
murcia_sf <- mapSpain::esp_get_ccaa("Murcia") |> 
  sf::st_transform(25830)
portugal_sf <- worldmap_sf |> 
  dplyr::filter(geounit == "Portugal")

species_name_italics <- expression(italic("Brachycaudus schwatzi"))

aphid_risk_map +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(title = "Risk Map for Andalucía",
       subtitle = species_name_italics,
       caption = "Climate data from WorldClim Historical data set",
       fill = "Annual Risk Months") +
  theme(legend.position = "bottom") +
  theme(plot.title = element_text(face = "bold")) +
  geom_sf(data = extremadura_sf, fill = "gray85",
          color = NA) +
  geom_sf(data = clm_sf, fill = "gray85",
          color = NA) +
  geom_sf(data = murcia_sf, fill = "gray85",
          color = NA) +
  geom_sf(data = portugal_sf, fill = "gray85",
          color = NA) +
  geom_sf(data = andalucia_sf, 
          color = "#060d0e",
          fill = NA) +
  geom_sf(data = bath_spain,
          fill = "#060d0e",
          color = "#060d0e")
```

Similarly, a discrete *"iso-risk"* map can be generated using `terra::geom_spatraster_contour_filled()` with the previous approach as follows:

```{r}
breaks_levels <- purrr::map_chr(.x = seq(0, max_risk_andalucia),
                                .f = ~paste0("[", .x, " ", .x+1,")")) |> 
  forcats::as_factor()

ggplot2::ggplot() +
  ggplot2::geom_sf(data = worldmap_sf, # <- countries background
                   fill = "gray85") +
  ggplot2::lims(x = c(bbox_rast[1], bbox_rast[3]),
                y = c(bbox_rast[2], bbox_rast[4])) +
  tidyterra::geom_spatraster_contour_filled(data = risk_rast_andalucia_epsg25830,
                                            breaks = seq(0, 12),
                                            maxcell = Inf) +
  ggplot2::geom_sf(data = worldmap_sf,
                   fill = NA,
                   color = "lightblue4") +
  ggplot2::scale_fill_manual(values = my_risk_palette,
                             labels = breaks_levels) +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(title = "Risk Map for Andalucía",
       subtitle = species_name_italics,
       caption = "Climate data from WorldClim Historical data set",
       fill = "Annual Risk Months") +
  theme(legend.position = "bottom") +
  theme(plot.title = element_text(face = "bold")) +
  geom_sf(data = extremadura_sf, fill = "gray85",
          color = NA) +
  geom_sf(data = clm_sf, fill = "gray85",
          color = NA) +
  geom_sf(data = murcia_sf, fill = "gray85",
          color = NA) +
  geom_sf(data = portugal_sf, fill = "gray85",
          color = NA) +
  geom_sf(data = andalucia_sf, 
          color = "#060d0e",
          fill = NA) +
  geom_sf(data = bath_spain,
          fill = "#060d0e",
          color = "#060d0e")
```

## References
