---
title: "Generating Risk Maps"
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mappestRisk)
```

## Generating risk maps for pest development

### Overall approach and ecological meaning

`mappestRisk` aims to provide an easy-to-use modelling workflow for producing pest risk maps.

Development rates variation across temperatures has been largely used to produce risk maps of pest occurrence, usually maps of annual generations for arthropod crop pests based on linear models (see e.g., [@efsa2017]). Despite more complex approaches have been recently used for pest risk assessment based on biological responses to temperature[@efsa2023], nonlinear TPC modelling of development rates variation across temperatures, already available after decades of experimental data collection, can help inform pest risk assessment throughout multiple crop pests[^1].

[^1]: For theoretical insights on the relationship between development rate TPCs and other population growth rates approximating fitness, see [@pawar2024; @amarasekare2012].

The approach followed by `mappestRisk` is inspired by the one developed by @taylor2019. These researchers compose a mathematical model through TPC model fiting that describes the temperature-dependence of huanglongbing transmission from *Diaphorina citri* to citrus plants. Given that this model theoretically approximates the fundamental thermal niche of the species, they project throughout citrus growing regions how many months per year have average temperatures that result optimal for transmitting the virus. This index (number of high risk months) results extremely helpful for policy making, since it reduces sources of uncertainty in the communication field by using a direct, applied measurement [@simmonds2022].

Given the major contribution of development rate TPCs to ectotherms' fitness[@pawar2024; @amarasekare2012] and thus likely on fundamental thermal niche, here we suggest applying a similar framework in a simpler way to calculate how many months per year are optimal for a population to maximize its development.

### Calculate the Thermal Suitability Boundaries

In order to obtain thermal suitability boundaries or limits defining the TPC region where development is maximum, the package incorporates the `therm_suit_bounds()` function.

This function has as input the `tibble` obtained after using the `predict_curves()` function, either with- or without uncertainty propagation. Note that only one model can be used for calculating thermal suitability boundaries at a time. Additionally, it has a `suitability_threshold` argument to allow the user to specify which quantile of the TPC is being used. In other words, a `suitability_threshold = .75` will calculate the two temperature values at which the TPC predicts a value $R(T_{75}) = 0.75\times\ r_\max$ , one at each side of the TPC peak. If multiple TPCs are given due to setting the `predict_curves()` argument `propagate_uncertainty` to `TRUE`, two boundaries will be calculated for each TPC. This function outputs a `tibble` with the thermal suitability boundaries for each curve.

Note that setting `suitability_threshold` to `0` will yield the thermal limits of the complete TPC, i.e., the $CT_\min$ and the $CT_\max$. While the map resulting from default (.75) or similar threshold measures indicates severe risk (populations may rapidly develop), maps from permissive measures such as critical thermal limits may rather indicate risk of the pest to find a refuge in the selected region.

Here we provide a example[^2]:

[^2]: Warnings have been omitted to avoid overwhelming information in the article, but they usually refer to TPC simulations yielding `NA` values for thermal suitability boundaries (see "chopped" simulation curves in the `plot_uncertainty()` output in [Simulate TPCs with bootstrap to propagate uncertainty](articles/tpcs-simulate-bootstrap.html)article.

```{r, message=FALSE, warning=FALSE}
#fit previously:
data("aphid")
fitted_tpcs_aphid <- fit_devmodels(temp = aphid$temperature,
                                   dev_rate = aphid$rate_value,
                                   model_name = "all")
#predict curves previously
preds_boots_aphid <-predict_curves(temp = aphid$temperature,          
                                   dev_rate = aphid$rate_value,
                                   fitted_parameters = fitted_tpcs_aphid,
                                   model_name_2boot = c("briere2", "lactin2"),
                                   propagate_uncertainty = TRUE,
                                   n_boots_samples = 100)
```

```{r, warning=FALSE, message=FALSE}

boundaries_aphid <- therm_suit_bounds(preds_tbl = preds_boots_aphid,       
                                      model_name = "lactin2",        
                                      suitability_threshold = 80) 
```

The output has the following aspect:

```{r}
print(boundaries_aphid)
```

### Project a risk map

Given the fitted TPCs and their calculated suitability thermal traits, `mappestRisk` enables to automatically generate rasters –`terra::SpatRaster`– at the targeting region to forecast summarizing the number of months per year with highly suitable temperatures for pest development.

This can be done with the function `map_risk()`

The workflow of the function is as follows:

1.  **Extract temperature data:** When no temperature raster is provided by the user in the `t_rast` argument (default `NULL`), `map_risk()` automatically downloads temperature data from [WorldClim](https://worldclim.org/data/worldclim21.html) historical data set using `geodata` [@hijmans2024] package, consisting in average temperature at monthly time resolution and a user-defined spatial resolution through `res` argument –our function uses 2.5 resolution or \~4.625km at the equator by default. If the `region` argument is a string with any country name available in `data(country_names)`, the function calls `geodata::worldclim_country()`. Alternatively, if the `region` argument is either a `SpatExtent`, a vector with the numeric boundaries of this spatial extent or a `sf` or `SpatVector` object –i.e., polygons, the `map_risk()` function will download data for the entire world using `geodata::worldclim_global()`[^3].
2.  **Spatial operations**: Once the temperature raster is available (either through user-input in `t_rast` or by downloading from WorldClim), the `map_risk()` function operates using `terra` R package for spatial operations such as cropping the data to the target region spatial extent and masking (if `mask = TRUE`). Next, `map_risk()` uses the input of the argument `t_vals` –which has to be a `tibble` exported from `therm_suit_bounds()`– to assign a value of `1` at those cells with monthly temperatures within the interval given by `therm_suit_bounds()` or `0` if not, yielding a raster with 12 layers (1 for each month) with `0`'s and `1`'s. Then, it automatically sums those values for each cell across the 12 layers to give a final `SpatRaster` with one layer whose values vary between 0 or 12, indicating the number of months per year with highly suitable temperatures –within the thermal boundaries from `therm_suit_bounds()` from TPC model fitting– at that cell, or in other words, the *risk value*. When more than one row is given in `t_vals` (due to multiple thermal boundaries after propagating uncertainty with `predict_curves()`), one "partial" `SpatRaster` is calculated by `map_risk()` for each pair of thermal boundaries. In this case, a final `SpatRaster` is generated with two values layers: the *Risk* layer that averages the *risk* value across partial rasters and the *uncertainty* raster that calculates the standard deviation across these partial rasters as an uncertainty risk measurement.
3.  **Visualization**: despite the function yields a `SpatRaster`, it automatically plots the map representing the raster vallues whenever the `plot` argument is set to `TRUE`. This map is generated using `terra::plot()` and will have one panel showing the risk value –i.e., the *risk* raster layer– and, when parameter uncertainty has been included in the complete `mappestRisk` workflow, it will add a second panel showing the standard deviation of the risk value in `acton` palette –i.e., the *uncertainty* raster layer. Cells are coloured based on their values using *bilbao* and *acton* palettes from `khroma` package [@khroma].
4.  **Interactive maps:** `map_risk()` has an additional logical argument called `interactive`, default to `FALSE`. If set to `TRUE`, the output of `map_risk` won't be a `SpatRaster` any longer but a `leaflet` object instead, and an interactive map with the above-mentioned layers will be generated using `terra::plet()`.

[^3]: We recommend to think carefully where to extract data since a simple call to `map_risk()` without a country name will download \~450MB of climate data with default conditions (`res = 2.5`). Note that the `path` argument must be given by the user. If the user desires to save the temperature data sourcing the map, this path should refer to somewhere in your device; otherwise, a `tempdir()` is recommended to avoid large data storage.

The following sections show how to obtain risk rasters with default visualization given directly by `map_risk()`. Custom alternatives for map visualization are detailed in [Customize your Risk Maps](articles/custom-risk-maps.html) article.

### Risk map: country

If we follow the workflow of the package in the examples elsewhere in this website, we can then create a risk map for a country. In this example, we will obtain it for Taiwan.

Since '*Taiwan'* is the official name included in `mappestRisk::country_names` and no `t_rast` is given by the user, the function will automatically download data cropped for this country.

```{r, message = FALSE, warning=FALSE, echo=FALSE}

# STEP 1: Fit TPCs
data("aphid")
fitted_tpcs_aphid <- fit_devmodels(temp = aphid$temperature,
                                   dev_rate = aphid$rate_value,
                                   model_name = "all")
## optional: show predictions

# plot_devmodels(temp = aphid$temperature,
#                dev_rate = aphid$rate_value,
#                fitted_parameters = fitted_tpcs_aphid,
#                species = "Brachycaudus schwartzi",
#                life_stage = "nymphs")

# STEP 2: Obtain predictions and bootstrapped TPCs

preds_boots_aphid <- predict_curves(temp = aphid$temperature,          
                                   dev_rate = aphid$rate_value,
                                   fitted_parameters = fitted_tpcs_aphid,
                                   model_name_2boot = c("briere2", "lactin2"),
                                   propagate_uncertainty = TRUE,
                                   n_boots_samples = 100)
# optional: show simulated curves:
# plot_uncertainties(bootstrap_uncertainties_tpcs = preds_boots_aphid,
#                    temp = aphid$temperature,
#                    dev_rate = aphid$rate_value,
#                    species = "Brachycaudus schwartzi",
#                    life_stage = "Nymphs")

# STEP 3: Calculate thermal suitability boundaries

boundaries_aphid <- therm_suit_bounds(preds_tbl = preds_boots_aphid,       
                                      model_name = "lactin2",        
                                      suitability_threshold = 80) 
print(boundaries_aphid)
```

```{r}

# Create a Risk Map for Taiwan

risk_rast_taiwan <- map_risk(t_vals = boundaries_aphid, 
                             path = tempdir(), # directory to download data 
                             region = "Taiwan",    
                             mask = TRUE,
                             plot = TRUE,
                             interactive = FALSE,
                             verbose = TRUE)
```

The map on the left side shows how many months have optimal temperatures for the aphid *Brachycaudus schwartzi* along Taiwan, from 0 (in mountains and highlands) up to 9 (in southern lowlands). The map on the right side shows an error map from parameter uncertainties, with regions at which the risk estimate is exact (e.g., the mountains) and others with up to 1.6 yielding the standard deviation of the risk value (in *number of suitable months*) at specific cells.

### Risk map: Spatial Extent

In the following example, the `region` will be a frame with four coordinates (`xmin, xmax, ymin, ymax`) encompassing the Canary Islands in WGS84 (or EPSG:4326) projection that is use by default in `map_risk()` function. First, let's convert the numeric extent into a `SpatExtent` using `terra::ext()`[^4]:

[^4]: Note that `map_risk()` will also work for a numeric vector with this four values, so using `terra::ext()` is not neccesary and we use it for safety practices writing (it might avoid typo data, etc).

```{r}
canary_islands_spatextent <- terra::ext(c(-19, -13, 27.5, 29.5))
```

Then, we applied the same function as before for the `aphid` workflow:

```{r}
risk_rast_canary <- map_risk(t_vals = boundaries_aphid, 
                             path = tempdir(), # directory to download data 
                             region = canary_islands_spatextent,    
                             mask = TRUE,
                             plot = TRUE,
                             interactive = FALSE,
                             verbose = TRUE)
```

### Risk map: Polygons

A typical scenario may involve forecasting data to regional or province level, to natural protected areas or to crop's growing regions. In these cases, the user input for `region` must be an `sf` object –e.g., by loading a shapefile (\**.shp)* in R using `sf::st_read()`– or as a `SpatVector` –e.g., by loading a shapefile using `terra::vect()`. The `map_risk()` function will automatically download global data and crop it to the spatial extent of the object, and will perform a masking operation if `mask = TRUE`.

In the following example, we will use it for Andalucía region in mainland Spain. The polygon will be obtained using `mapSpain` package [@mapSpain], which uses "ETRS89" as default coordinates reference system. The `map_risk()` function will automatically transform it for the climate extraction and the procedure performs similar to the previous examples:

```{r}
andalucia_sf <- mapSpain::esp_get_ccaa(ccaa = "Andalucía")

risk_rast_andalucia <- map_risk(t_vals = boundaries_aphid, 
                                path = tempdir(), # directory to download data 
                                region = andalucia_sf,    
                                mask = TRUE,
                                plot = TRUE,
                                interactive = FALSE,
                                verbose = TRUE)
```

## References
