---
title: "Interactive visualization"
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = "#>"
)
```

# Introduction

In this vignette we demonstrate how to generate interactive maps from the results obtained using the main workflow of the **mappestRisk** package. The **mappestRisk** package provides tools to estimate climatic suitability and developmental risk for ectothermic organisms based on temperature-dependent development models. The main workflow produces spatial outputs summarizing the number of months in which environmental conditions allow development.

These results could be returned as a multi-layer raster object containing:

- the **mean number of suitable months** (*risk*), and  
- the **uncertainty associated with predictions** (standard deviation across bootstrap simulations).

While static maps are useful for reporting results, interactive visualization allows users to explore spatial patterns more dynamically and intuitively. In this article we demonstrate how to build an **interactive map** from the raster output produced by `map_risk()` using the **leaflet** ecosystem in R.

# Setup

```{r setup}
library(mappestRisk)
library(leaflet)
library(leaflet.opacity)
library(leafem)

library(terra)
library(khroma)
library(htmlwidgets)
```

# Obtaining the risk raster

To demostrate how to use the interactiviy we use the main workflow of the **mappestRisk** package (see [here](../../README.Rmd)). The raster object required for the interactive map is produced by the `mappestRisk::map_risk()` function. The typical steps of the mappestRisk workflow include: 

1. fitting temperature–development models  
2. propagating uncertainty through bootstrap simulations  
3. estimating thermal suitability boundaries  
4. projecting these limits into geographic space


```{r, echo = FALSE, fig.show = 'hide'}

data("aphid") 
fitted_tpcs <- fit_devmodels(temp = aphid$temperature,  
                             dev_rate = aphid$rate_value, 
                             model_name = c("briere2", "lactin2", "thomas"))

plot_devmodels(temp = aphid$temperature,
               dev_rate = aphid$rate_value,
               fitted_parameters = fitted_tpcs,
               species = "Brachycaudus schwartzi",
               life_stage = "Nymphs")

preds_boots <- predict_curves(temp = aphid$temperature,          
                              dev_rate = aphid$rate_value,
                              fitted_parameters = fitted_tpcs,
                              model_name_2boot = c("briere2", "lactin2", "thomas"),
                              propagate_uncertainty = TRUE,
                              n_boots_samples = 10)

boundaries <- therm_suit_bounds(preds_tbl = preds_boots,
                                model_name = "lactin2",        
                                suitability_threshold = 80) 



risk_rast <- map_risk(t_vals = boundaries, 
                      region = "Réunion",   
                      path = tempdir() # directory to download climate data 
)
```

# Preparing the layers

- First, we obtained the layers and create several custom palettes. 

```{r palettes, }
risk <- risk_rast[["mean"]]
uncertainty <- risk_rast[["sd"]]

risk_colors <- c(
  "white",
  grDevices::colorRampPalette(
    c("#0081a7","#00afb9","#fdfcdc","#fed9b7","#f07167")
  )(12)
)

pal_risk <- leaflet::colorNumeric(
  palette = risk_colors,
  domain = c(0,13),
  na.color = "transparent"
)

uncertainty_colors <- as.character(khroma::colour("bilbao")(100))
mm <- terra::minmax(uncertainty)

pal_uncertainty <- leaflet::colorNumeric(
  palette  = uncertainty_colors,
  domain   = c(mm[1,1], mm[2,1]),
  na.color = "transparent"
)
```


# Building the custom interactive map

The interactive map includes several features that improve usability and facilitate data exploration:

- **Mini map**. A small overview map can be added using `leaflet::addMiniMap()` [@Rleaflet] to help users orient themselves while navigating the map.

- **Opacity control**. The **leaflet.opacity** package [@Rleafletopacity] provides `leaflet.opacity::addOpacitySlider()`, allowing users to interactively adjust the transparency of the raster layer.

- **Pixel value query**. With `leafem::addImageQuery()` [@Rleafem], the value of the raster cell is displayed as the cursor moves over the map, enabling quick inspection of local predictions.


```{r leaflet-map}
m <- leaflet() |>
  addProviderTiles(providers$CartoDB.Positron, group = "Basemap") |>
  addProviderTiles(providers$Esri.WorldImagery, group = "Satellite") |>
  addRasterImage(
    risk,
    layerId = "Risk Map",
    group = "Risk Map",
    colors = pal_risk
  ) |>
  addRasterImage(
    uncertainty,
    layerId = "Uncertainty",
    group = "Uncertainty",
    colors = pal_uncertainty
  ) |>
  addMiniMap(toggleDisplay = TRUE) |>
  leaflet.opacity::addOpacitySlider(layerId = "Risk Map") |>
  leaflet.opacity::addOpacitySlider(layerId = "Uncertainty") |>
  addLayersControl(
    position = "bottomright",
    baseGroups = c("Satellite", "Basemap"),
    overlayGroups = c("Risk Map", "Uncertainty"),
    options = layersControlOptions(collapsed = TRUE)
  ) |>
  leafem::addImageQuery(
    risk,
    project = FALSE,
    type = "mousemove",
    layerId = "Risk Map",
    digits = 2
  ) |>
  leafem::addImageQuery(
    uncertainty,
    project = FALSE,
    type = "mousemove",
    layerId = "Uncertainty",
    digits = 2
  ) |>
  addLegend(
    pal = pal_risk,
    values = c(0,3,6,9,12),
    title = "Risk Maps<br>(n months)",
    group = "Risk Map",
    position = "bottomleft",
    labFormat = leaflet::labelFormat(digits = 0)
  ) |>
  addLegend(
    pal = pal_uncertainty,
    values = c(mm[1,1], mm[2,1]),
    title = "Uncertainty<br>(n months)",
    group = "Uncertainty",
    position = "bottomleft"
  ) |>
  hideGroup("Uncertainty")

m
```

# Exporting the map

Leaflet maps created in R are HTML widgets, which means they can be easily exported as standalone HTML files and shared with others. This allows the interactive map to be opened in any web browser without requiring an R environment.

The function `htmlwidgets::saveWidget()` [@Rhtmlwidgets] can be used to save the map as a self-contained HTML file that includes all necessary dependencies.

```{r export}
outfile <- file.path(tempdir(), "interactive_map.html")

htmlwidgets::saveWidget(
  widget = m,
  file = outfile,
  selfcontained = TRUE
)
```

# Summary

Interactive maps provide an effective way to explore and communicate spatial risk estimates produced by **mappestRisk**.

Combining raster outputs with **leaflet** allows users to inspect spatial patterns, evaluate uncertainty, and share interactive results easily.

# References 

