---
title: "TPCs model fitting"
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include = FALSE}
library(mappestRisk)
```

## Fit a thermal performance curve (TPC) to your data:

The Thermal Performance Curves can be understood as models describing the nonlinear temperature dependence of several biological rates [@amarasekare2012]. TPC shapes are usually unimodal and left-skewed [@huey1979], which is also the case for development rate TPCs of insects [@régnier2022]. Fitting these nonlinear regression models requires to incorporate *a priori* knowledge on model parameters. For that purpose, the R packages `rTPC` [@padfield2021; @rTPC]and `devRate`[@devRate; @rebaudo2018] have provided R functions and workflows to facilitate model fitting, start values selection and parameter calculation of TPCs. Additionally, `nls.multstart` package [@padfield2021; @nls.multstart] provides a nonlinear, least-squares framework to facilitate fitting these curves to experimental data in a flexible and efficient manner.

However, we believe that most researchers obtaining experimental data on rate development across temperatures may not have enough R programming knowledge to accomplish repeteadly the model fitting procedure in reproducible workflows. Aiming to facilitate the usage of these previous contributions made by the above-mentioned R packages to TPC modelling for a broad-audience among experimental researchers, `mappestRisk` aims to synthesize these software development into an unique function called `fit_devmodels()`.

This function takes for input a `data.frame` or `tibble` containing two numeric columns: one with temperature data –with at least 4 unique temperature treatments– and the other with development rate data. The user can select which models to fit among those included in `available_models` (or use them all). Then, for each model equation, the function will search for the most appropriate start values data for performing nonlinear regression models using `rTPC::get_start_vals()` when possible, or alternatively `devRate` parameter data set for models not included in the former package. Once selected, the function will also search for lower and upper thermal limits when possible using `rTPC::get_lower_lims()` or `rTPC::get_upper_lims()`. Next, the function performs sequentially the nonlinear least-squares regression models using `nls.multstart::nls_multstart()` function for each set of model equation, start values and limits with default configuration of `nls_multstart` approach (see function documentation here). TPC models not achieving convergence are further discarded.

In this example, we'll show how to fit one to several thermal performance curves to a data set of development rate variation across temperatures[^1]. The following code provides an example as given in `fit_devmodels()` function documentation, with a data table showing the output of fitted models.

[^1]: At least 4 unique temperatures are required. Fore more details, see documentation.

```{r,  warning=FALSE}
data("aphid")

fitted_tpcs_aphid <- fit_devmodels(temp = aphid$temperature,
                                   dev_rate = aphid$rate_value,
                                   model_name = "all")

```

After running `fit_devmodels()`, a `tibble` (a specific case of `data.frame`) is created containing the most relevant statistical information of the modelled TPCs: model name, parameter information –name, estimates, standard error, AICs and listed model objects.To continue the package workflow, this output must be stored as an object (here we call it `fitted_tpcs_aphid`). This `tibble` has the following aspect:

```{r}
print(fitted_tpcs_aphid)
```

Despite the output yields a table with explicit statistical information and model objects with implicit additional information, selecting a model should better be guided by ecological, rather that only statistical criteria. This typically involves visualising the predictions of the fitted TPCs for further visual examination. A potential user of the package may use their expertise on thermal biology and ecology of the species to distinguish which of the fitted TPCs accomplish to approximate true temperature-development relationships. For example, a fitted TPC model yielding a critical thermal maximum beyond 50ºC –i.e., with a right-side vertical x-axis cutoff– may not be realistic, as most species do not encounter this temperatures. Similarly, models yielding quasi-symmetric TPCs or unexpected behaviors at some regions of the curve may also be further discarded. We strongly believe there is no a straightforward statistical strategy to select among TPC models; instead, comparing a large number of fitted TPCs with expertise ecological criteria may improve biological realism of for less arbitrary model selection. Finally, accounting for the thermal regimes that the species will likely encounter in the region of interest to project risk may help select among curves by prioritizing curves with specific expected behaviors in the TPC regions corresponding to these thermal regimes.

The `plot_devmodels()` function facilitates quick visualizations of predictions of the Thermal Performance Curves to guide model selection under ecological criteria. It takes as an input the resulting `tibble`/`data.frame` stored as an object from `fit_devmodels()`. It has also arguments to title and subtitle the figure with the species' name and the life stage evaluated, respectively. Since wrapped in the `ggplot2` approach, customization is available by adding new lines with compatible functions with `ggplot2`.

Following the previous example:

```{r, fig.width=10, fig.height=8, dpi=300, out.width="100%"}
plot_devmodels(temp = aphid$temperature,
               dev_rate = aphid$rate_value,
               fitted_parameters = fitted_tpcs_aphid,
               species = "Brachycaudus schwartzi",
               life_stage = "nymphs")
```

As you can see, the plot is faceted across TPC models included in `fitted_tpcs_aphid`.

## References
