---
title: "Simulate TPCs with bootstrap to propagate uncertainty"
bibliography: references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(mappestRisk)
```

## Simulate $n$-thermal performance curves using bootstrap with residual resampling

### Why propagating uncertainty?

Forecasting necessarily incorporates uncertainties in how much we know about the knowledge of the target system (i.e., model structure, parameter uncertainty, response or predictor variable errors, etc) that adds together with uncertainty in how to communicate findings and an unavoidable randomness within natural systems [@simmonds2022].

At least three different uncertainty sources of the models should be addressed in forecasts:

1.  **Parameter uncertainty:** the accuracy of estimated parameters of the models may affect the confidence of the predictions. For example in the [TPCs model fitting](articles/tpcs-model-fitting.html) article, the `briere2` model yields $CT_\max=36.5\pm3.40 \ \textrm{ºC}$. Let's imagine a forecaster aiming to identify "safe" regions where the pest may not be established due to extremely high maximum temperatures (e.g., $T_\max > \bar{CT}_\max$). It's possible that all the forecasting regions have monthly maximum temperatures of about 34ºC, which lie below the $CT_\max$ estimate of 36.5 ºC, leading to identify no risk for the assessment. However, if we incorporate how the uncertainty of each parameter contributes to the variability of the predicted TPC with simulated TPC ribbons in the plot –see e.g., below with `plot_uncertainties()`, there are possible scenarios at which several TPC-calculated $CT_\max$'s lie below 34ºC, yielding a not-negligible risk likelihood.
2.  **Predictor uncertainty:** incorporating the variability of the predictor –in the above case, monthly maximum temperatures– will also yield a probability distribution of forecast outcomes (let's say $T_\max = 34 \textrm{ºC}, \ CI =[31.2, 36.8]$). This would result in some scenarios with maximum temperatures above $CT_\max$ estimate of $36.5$ºC and some others where they have not.
3.  **Source data uncertainty:** additionally, TPCs are usually fitted to summarized data from experiments in laboratory conditions. These measures incorporate both both mesasurement error (at the individual level) and uncertainty measures summarizing the variability of rate estimates at the population level.

For now, `mappestRisk` enables to explicitly account for parameter uncertainty by simulating $n$ TPCs using bootstrapping techniques with residual resampling, as suggested and implemented by `rTPC` package (see [this vignette/article](https://padpadpadpad.github.io/rTPC/articles/bootstrapping_many_curves.html) [@rTPC]) as described in the section below. Measurement uncertainty of source data might be further incorporated in future enhancement updates of the package through a `weights` argument of `fit_devmodels()` and `predict_curves()` [@padfield2021] since they are based on the `rTPC` - `nls.multstart` framework that has recently incorporated an [article on how to simulate curves by weighted bootstrapping](https://padpadpadpad.github.io/rTPC/articles/weighted_bootstrapping.html). Finally, a discussion on how to deal with predictor uncertainty of the forecasts and how to overcome communication uncertainty is given in the [generate-risk-maps vignette article](articles/generate-risk-maps.html).

### Simulate bootstrapped TPCs with `predict_curves()`

As mentioned above, `mappestRisk` includes two functions that automate the workflow suggested by `rTPC` package to simulate curves for propagating parameter uncertainty. We implemented the *residual resampling* method from `car::Boot()` following `rTPC` vignete suggestions, since the predictor data are controlled by the experimental researchers and usually few data points per curve with low representation of the hot decay TPC region in the source data. For further insights on the methodology, please refer to the [the original `rTPC` vignette](https://padpadpadpad.github.io/rTPC/articles/bootstrapping_models.html#residual-resampling). Further updates of the package may incorporate variance modelling for heteroskedastic residuals, as well as an argument to choice between *residual* *resampling* and *case resampling*[^1].

[^1]: More methods to perform these bootstrap resampling tecniques are given in [@sonderegger2020].

The `predict_curves()` function incorporates the temperature and development rate data arguments, a `fitted_parameters` argument that requires as input the table obtained from `fit_devmodels()`, a `model_name_2boot` argument to choose which TPC model(s) to bootstrap along those available in `fitted_parameters` and the number of bootstrap samples, or `n_boots_samples`. The function implicitly fits a new TPC for each model equation using `minpack.lm::nlsLM()` using the original parameter estimates –coefficients– from `fit_devmodels()` as start values. Then, the `car::Boot()` [^2]function with the arguments `method = "residual"` and `R = n_boots_samples` resamples the original data by adding the resampled residuals from `minpack.lm::nlsLM()` to the resampled fitted values and repeats the operation $n-$times –i.e., `n_boots_samples` times. Finally, these bootstrapped curves are predicted along temperature data –more specifically, 20ºC below and 15ºC above the minimum and maximum temperature values, respectively, in `temp` argument.

[^2]: Please, note that `car::Boot()` requires two specific conditions to work: (1) to have variables loaded in the local environment,

Here we have an example:

```{r, message = FALSE}
#fit previously:
data("aphid")
fitted_tpcs_aphid <- fit_devmodels(temp = aphid$temperature,
                                   dev_rate = aphid$rate_value,
                                   model_name = "all")
```

```{r}
preds_boots_aphid <-predict_curves(temp = aphid$temperature,          
                                   dev_rate = aphid$rate_value,
                                   fitted_parameters = fitted_tpcs_aphid,
                                   model_name_2boot = c("briere2", "lactin2"),
                                   propagate_uncertainty = TRUE,
                                   n_boots_samples = 100)
```

By default, the `predict_curves` function does propagate uncertainty by simulating as many curves as asked through `n_boots_samples` argument for each of the selected models from `model_name_2boot`. If able to perform bootstrap, this default configuration will output a `tibble` with simulated TPCs for both ploting purposes and thermal traits calculation. `n_boots_samples` is set up to 100 by default. We recommend to avoid lower values that may inaccurately reflect uncertainty and to think carefully when trying to perform bootstrap with larger samples, since they may exponentially increase computational demands with little benefit for the purposes addressed here. If `propagate_uncertainty` is set to `FALSE`, the function will output the same `tibble` with one single curve (temperatures and predictions) coming from the estimated TPC (similar to those plotted in `plot_devmodels()`for applying subsequent steps in the `mappestRisk` suggested workflow.

The `tibble` output of `predict_curves` can be easily visualized with `plot_uncertainties()`, where the curve from the model estimate is plotted as a thicker, dark orange line and the bootstrapped curves are depicted as slighter, dark blue lines composing a sort of a ribbon for the central curve. If more than one model has been succesfully bootstrapped, predicted curves will be plotted along different facets.

```{r}
plot_uncertainties(bootstrap_uncertainties_tpcs = preds_boots_aphid,
                   temp = aphid$temperature,
                   dev_rate = aphid$rate_value,
                   species = "Brachycaudus schwartzi",
                   life_stage = "Nymphs")
```

These simulated TPCs may guide ecologically realistic model selection and propagating parameter uncertainty for subsequent analyses. However, as for `plot_devmodels()`, we discourage selecting among TPC models solely based on statistical information, but rather on informed ecological criteria.

Please, consider carefully whether your data is suitable for these procedures.

## References
